Diverse Genotype-by-Weather Interactions in Switchgrass

### 2024-06-12 manuscript/revision2/revision2.qmd
revision2: generate revised figures 1-4 and SI figures 1-4

Narrow-sense heritabilities (h^2^) suggested that rank-changing GxE for these phenotypes was present across the common gardens (@fig-map C). h^2^ were typically high at individual gardens: 59% on average for green-up date, and 87% for flowering date. However, h^2^ were variable across gardens, and green-up dates were uncorrelated (r^2^ \< 0.2) or negatively correlated between pairs of gardens (@fig-map B). These negative and small correlations undoubtedly contributed to the low h^2^ values for green-up and flowering date when estimated jointly at all eight gardens: h^2^ was 0.8% for green-up and 23.2% for flowering date.

@@

We expected patterns of GxE and GxWeather to vary at the locus, subpopulation, and trait level; thus we were interested in modeling the different ways effects might covary across our eight common gardens. We jointly re-estimated genetic effects of a subset of SNPs across all eight common gardens using *mash* models, which can flexibly capture many modes of covariance across gardens (@fig-map D; SI Appendix, Sections S1-S5, Datasets 1-6). To obtain genetic effects, we used subsets of effects from garden- and subpopulation GWAS (SI Appendix, Section S2-S3; Table S1). To obtain covariance structures, we specified both covariance structures introduced in the initial *mash* manuscript, and GxWeather covariance matrices estimated from the covariance of empirical weather patterns at each garden at specific times before the phenological event (Table 1; SI Appendix, Section S1). Then, we used a greedy algorithm to select covariance matrices from each model's set that significantly improved the model likelihood (SI Appendix, Section S4; Table S2). Finally, we fit *mash* models using the selected set of covariance matrices on a subset relatively unlinked (r^2^ \< 0.2), 'strong' genetic effects with low *p*-values (SI Appendix, Section S5).

@@

Eight GxWeather covariance structures were selected by the greedy algorithm, one to two per subpopulation: five for green-up date, and three for flowering date (SI Appendix, Table S2). Of these eight, five had mass on them in *mash* models of the strong effects (@fig-covar A-B). Three of these five matrices had negative covariances between sites between Texas and North gardens (@fig-covar A), while two had all positive or near-zero covariances.

@@

For green-up date for the Gulf subpopulation, hundreds to thousands of pairwise effects exhibited rank-changing GxE, or a difference in sign between pairs of Texas and North gardens (@fig-effects B; SI Appendix, Fig S1). 78.7% of pairwise comparisons between North and Texas gardens had a difference in sign, while only 28.6% and 0.2% of North-North or Texas-Texas comparisons had a difference in sign, respectively (@fig-effects B). The majority of pairwise effects for greenup for the Midwest (\>55%) and Both (\>85%) subpopulations were the same sign, and effects most often differed in magnitude between gardens within and between the regions (@fig-effects B).

For flowering date for the Gulf subpopulation, less than 2% of pairwise effects exhibited rank-changing GxE, or a difference in effect sign within or between regions (@fig-effects B). More effects differed in magnitude between the Texas and North regions than within these regions (42.7% vs \<20%; @fig-effects B), while the majority of effects had no GxE. The Midwest population had relatively few significant effects for flowering, but a large proportion of these differed in sign between Texas and North regions (42.7%) or within the North region (65.4%). Finally, in Both subpopulations, less than 20% of pairwise effects differed in sign (@fig-effects B). Most differences in sign were between TX1, the southernmost garden, and all other gardens (SI Appendix, Figure S1). Similarly, more effects that differed in magnitude included gardens in the Texas region (52.3-55.9%), and most effect pairs in the North region were not distinguishable (91.5%).

@@

We sought additional experimental support the occurrence and location of our significant SNP effect re-estimates using an independent pseudo-F2 mapping population created from Gulf & Midwest individuals and grown at the same sites (@fig-qtl A,B). We conducted quantitative trait loci (QTL) mapping of flowering as functions of five environmental cues that we also used as covariance matrices in *mash*, and identified eight QTL for flowering date, eight QTL for flowering as a function of day length change two days prior, one QTL for the start of vegetative growth, and two QTL for vegetative growth as a function of daylength change one day prior, all of which showed QTL by environment interactions (SI Appendix, Figure S2). All QTL for flowering overlapped one or more homologs from rice or *A. thaliana* with functionally validated roles in flowering (SI Appendix, Dataset 7); the QTL on Chr04K overlapped a gene recently functionally validated for flowering in switchgrass [@choi2023]. All flowering and green-up QTL intervals contained at least one SNP significant in at least one *mash* run at a log10-transformed Bayes Factor \> 2, or in the 1% tail of significance, whichever was stricter (SI Appendix, Dataset 8). We also looked for enrichments of *mash* SNPs in the 1% tail of significance (the '*mash* 1% tail') within each QTL interval. At the 5% level, three QTL had enrichments of SNPs in the *mash* 1% tail. Overall, there were five significant enrichments (p \< 0.05, hypergeometric test) of SNPs in the *mash* 1% tail in the QTL intervals. Thus, we were able to experimentally support the genomic windows of some re-estimates of significant SNP effects with a QTL mapping experiment using a separate mapping population.

@@

\end{table}
\clearpage
\twocolumn
```
![Characterization of the timing of vegetative (green-up) and reproductive (flowering) growth in the switchgrass diversity panel. (a) Map and trait histograms of green-up and flowering dates across two genetically distinct switchgrass subpopulations and eight common gardens. Purple represents individuals from the Midwest genetic subpopulation, and pink individuals from the Gulf subpopulation; map positions represent the original collection locations for the genotypes, and shapes represent the ecotype of the genotype. Vertical dashed lines indicate the summer solstice. Common gardens are arranged in latitudinal order. (b) Phenotypic correlations between clonal replicates planted at eight common gardens, within and between two genetic subpopulations. (c) Narrow sense heritability of green-up and flowering within single common gardens (purple) and across all eight common gardens (green), within and between two genetic subpopulations. (d) Flow diagram of the methods applied to the green-up and flowering dates to jointly estimate SNP effects across all sites. *Mash* was fit to SNP effect data and used to find covariance matrices that improved the *mash* model likelihood using a large set of randomly selected, relatively unlinked SNP effects; this model was applied to a "strong" set of SNP effects with large effect sizes in the univariate GWAS.](images/Fig1_200dpi.png){#fig-map}

![Example hypothesis-driven covariance matrices specified in *mash* and the posterior weights placed on all covariance matrices. (a) Left column: Five example hypothesized covariance matrices specified for the green-up date or flowering date phenotype; these matrices were created from environment-specific correlations across eight common gardens, and are described in Table 1. Common gardens are arranged in latitudinal order within the matrices. Right column: Five example canonical covariance matrices. Canonical matrices (purple) have simple interpretations, such as equal effects across all common gardens, or effects specific to a single common garden. (b,d) Total posterior weight placed on each covariance matrix type specified for (b) green-up date and (d) flowering date *mash* models, within and between two genetic subpopulations. Hypothesized covariance matrices (green). Covariance matrices included in *mash* that had zero posterior weight in all three *mash* runs on the genetic subpopulations, such as the identity matrix, are not shown. (c,e) Total posterior weight placed on covariance matrices that were hypothesized or canonical, for the (c) green-up date phenotype and (e) flowering date phenotype.](images/Figure_2_recolor_covariance_matrices_and_posterior_weights.png){#fig-covar}

![Types of GxE present between of pairs of jointly re-estimated effects in eight common gardens, for effects with lfsr \< 0.05. a) Examples of effect patterns at three pairs of sites with three types of GxE. All effects are for an alternate allele, with the reference allele effect defined to be zero at both gardens. Sign: Effects that differ in sign at these pairs of gardens (p \< 0.05, lfsr). Magnitude: Effects identical in sign (p \< 0.05, lfsr) that differ in magnitude by a factor of \>0.4. None: Effects not distinguishable by magnitude nor sign of the effect. b) The fraction of effects with each GxE type for the start of vegetative growth (Greenup Date) and reproductive growth (Flowering Date), within and between two genetic subpopulations. Common gardens are grouped by the larger region they came from: North gardens are within the natural range of the Midwest subpopulation, while Texas gardens are within the natural range of the Gulf subpopulation.](images/Figure_3_GxE.png){#fig-effects}

![Overlaps of QTL from an outbred pseudo-F2 cross and with jointly re-estimated SNP effects in the 1% tail of significance from a diversity panel. Dotted lines indicate permutation-based significance thresholds for each weather-related function. Stars indicate QTL with significant enrichment for SNPs in the 1% *mash* tail; G, M, and B indicate which subpopulation had enrichment: G - Gulf subpopulation, M -Midwest subpopulation, B - both subpopulations. Rug plots show genomic locations of SNPs in the 1% *mash* tail for flowering date for each subpopulation. a) QTL mapping for the start of vegetative growth (Greenup Date), and three weather-related functions of greenup date. b) QTL mapping for the start of reproductive growth (Flowering Date), and two weather-related functions of flowering date.](images/Figure_4_QTL_Overlaps.png){#fig-qtl}

>>>>

Narrow-sense heritabilities (h^2^) suggested that rank-changing GxE for these phenotypes was present across the common gardens (SI Appendix, Figure S1). h^2^ were typically high at individual gardens: 59% on average for green-up date, and 87% for flowering date. However, h^2^ were variable across gardens, and green-up dates were uncorrelated (r^2^ \< 0.2) or negatively correlated between pairs of gardens (@fig-map B). These negative and small correlations undoubtedly contributed to the low h^2^ values for green-up and flowering date when estimated jointly at all eight gardens: h^2^ was 0.8% for green-up and 23.2% for flowering date (SI Appendix, Figure S1).

@@

We expected patterns of GxE and GxWeather to vary at the locus, subpopulation, and trait level; thus we were interested in modeling the different ways effects might covary across our eight common gardens. We jointly re-estimated genetic effects of a subset of SNPs across all eight common gardens using *mash* models, which can flexibly capture many modes of covariance across gardens (SI Appendix, Sections S1-S5, Datasets 1-6). To obtain genetic effects, we used subsets of effects from garden- and subpopulation GWAS (SI Appendix, Section S2-S3; Table S1). To obtain covariance structures, we specified both covariance structures introduced in the initial *mash* manuscript, and GxWeather covariance matrices estimated from the covariance of empirical weather patterns at each garden at specific times before the phenological event (Table 1; SI Appendix, Section S1). Then, we used a greedy algorithm to select covariance matrices from each model's set that significantly improved the model likelihood (SI Appendix, Section S4; Table S2). Finally, we fit *mash* models using the selected set of covariance matrices on a subset relatively unlinked (r^2^ \< 0.2), 'strong' genetic effects with low *p*-values (SI Appendix, Section S5).

@@

Eight GxWeather covariance structures were selected by the greedy algorithm, one to two per subpopulation: five for green-up date, and three for flowering date (SI Appendix, Table S2). Of these eight, five had mass on them in *mash* models of the strong effects (@fig-covar B,D). Three of these five matrices had negative covariances between sites between Texas and North gardens (@fig-covar A), while two had all positive or near-zero covariances.

@@

For green-up date for the Gulf subpopulation, hundreds to thousands of pairwise effects exhibited rank-changing GxE, or a difference in sign between pairs of Texas and North gardens (@fig-effects B; SI Appendix, Fig S2). 78.7% of pairwise comparisons between North and Texas gardens had a difference in sign, while only 28.6% and 0.2% of North-North or Texas-Texas comparisons had a difference in sign, respectively (@fig-effects B). The majority of pairwise effects for greenup for the Midwest (\>55%) and Both (\>85%) subpopulations were the same sign, and effects most often differed in magnitude between gardens within and between the regions (@fig-effects B; SI Appendix, Fig S2).

For flowering date for the Gulf subpopulation, less than 2% of pairwise effects exhibited rank-changing GxE, or a difference in effect sign within or between regions (@fig-effects B). More effects differed in magnitude between the Texas and North regions than within these regions (42.7% vs \<20%; @fig-effects B), while the majority of effects had no GxE. The Midwest population had relatively few significant effects for flowering, but a large proportion of these differed in sign between Texas and North regions (42.7%) or within the North region (65.4%). Finally, in Both subpopulations, less than 20% of pairwise effects differed in sign (@fig-effects B). Most differences in sign were between TX1, the southernmost garden, and all other gardens (SI Appendix, Figure S3). Similarly, more effects that differed in magnitude included gardens in the Texas region (52.3-55.9%), and most effect pairs in the North region were not distinguishable (91.5%).

@@

We sought additional experimental support the occurrence and location of our significant SNP effect re-estimates using an independent pseudo-F2 mapping population created from Gulf & Midwest individuals and grown at the same sites (@fig-qtl A,B). We conducted quantitative trait loci (QTL) mapping of flowering as functions of five environmental cues that we also used as covariance matrices in *mash*, and identified eight QTL for flowering date, eight QTL for flowering as a function of day length change two days prior, one QTL for the start of vegetative growth, and two QTL for vegetative growth as a function of daylength change one day prior, all of which showed QTL by environment interactions (SI Appendix, Figure S4). All QTL for flowering overlapped one or more homologs from rice or *A. thaliana* with functionally validated roles in flowering (SI Appendix, Dataset 7); the QTL on Chr04K overlapped a gene recently functionally validated for flowering in switchgrass [@choi2023]. All flowering and green-up QTL intervals contained at least one SNP significant in at least one *mash* run at a log10-transformed Bayes Factor \> 2, or in the 1% tail of significance, whichever was stricter (SI Appendix, Dataset 8). We also looked for enrichments of *mash* SNPs in the 1% tail of significance (the '*mash* 1% tail') within each QTL interval. At the 5% level, three QTL had enrichments of SNPs in the *mash* 1% tail. Overall, there were five significant enrichments (p \< 0.05, hypergeometric test) of SNPs in the *mash* 1% tail in the QTL intervals. Thus, we were able to experimentally support the genomic windows of some re-estimates of significant SNP effects with a QTL mapping experiment using a separate mapping population.

@@

\end{table}
\clearpage
\twocolumn
```
![Characterization of the timing of the onset of vegetative (green-up) and reproductive (flowering) growth in the switchgrass diversity panel. (a) Map and trait histograms of green-up and flowering dates across two genetically distinct switchgrass subpopulations and eight common gardens. Purple represents individuals from the Midwest genetic subpopulation, and pink individuals from the Gulf subpopulation; map positions represent the original collection locations for the genotypes, and shapes represent the ecotype of the genotype. Histogram vertical dashed lines indicate the summer solstice. Common gardens are arranged in latitudinal order. (b) Phenotypic correlations between clonal replicates planted at eight common gardens, within and between two genetic subpopulations.](images/Figure_1_AB.png){#fig-map}

![Example hypothesis-driven covariance matrices specified in *mash* and the posterior weights placed on all covariance matrices. (a) Common gardens are arranged in latitudinal order within the matrices. Top row: Four example canonical covariance matrices. Canonical matrices (purple) have simple interpretations, such as equal effects across all common gardens, or effects specific to a single common garden. Bottom row: Five example GxWeather covariance matrices specified for the green-up date or flowering date phenotype; these matrices were created from environment-specific correlations across eight common gardens, and are described in Table 1. (b,d) Total posterior weight placed on each covariance matrix type specified for (b) green-up date and (d) flowering date *mash* models, within and between two genetic subpopulations. Covariance matrices included in *mash* that had zero posterior weight in all three *mash* runs on the genetic subpopulations, such as the identity matrix, are not shown. (c,e) Total posterior weight placed on covariance matrices that were hypothesized or canonical, for the (c) green-up date phenotype and (e) flowering date phenotype.](images/Figure_2.png){#fig-covar}

![Types of GxE present between of pairs of jointly re-estimated effects in eight common gardens, for effects with lfsr \< 0.05. a) Examples of effect patterns at three pairs of sites with three types of GxE. All effects are for an alternate allele, with the reference allele effect defined to be zero at both gardens and represented by the dashed vertical line. Sign: Effects that differ in sign at these pairs of gardens (p \< 0.05, lfsr). Magnitude: Effects identical in sign (p \< 0.05, lfsr) that differ in magnitude by a factor of \>0.4. Not Distinguishable: Effects not distinguishable by magnitude nor sign of the effect, with no measurable GxE. b) The fraction of effects with each GxE type for the onset of vegetative growth (green-up date) and reproductive growth (flowering date), within and between two genetic subpopulations. Common gardens are grouped by the larger region they came from: North gardens are within the natural range of the Midwest subpopulation, and include MO, NE, MI, and SD, while Texas gardens are within the natural range of the Gulf subpopulation, and include TX1, TX2, and TX3.](images/Figure_3_GxE.png){#fig-effects}

![Overlaps of QTL from an outbred pseudo-F2 cross and with jointly re-estimated SNP effects in the 1% tail of significance from a diversity panel. Dotted lines indicate permutation-based significance thresholds for each weather-related function. Stars indicate QTL with significant enrichment for SNPs in the 1% *mash* tail; G, M, and B indicate which subpopulation had enrichment: G - Gulf subpopulation, M -Midwest subpopulation, B - both subpopulations. Rug plots show genomic locations of SNPs in the 1% *mash* tail for flowering date for each subpopulation: Both subpopulations are above the plot panel, the Gulf subpopulation is above the x-axis, and the Midwest subpopulation is below the x-axis. a) QTL mapping for the onset of vegetative growth (green-up date), and three weather-related functions of green-up date. b) QTL mapping for the onset of reproductive growth (flowering date), and two weather-related functions of flowering date.](images/Figure_4_QTL_Overlaps.png){#fig-qtl}
